### 2017-08-28 reports/thesis/book.tex
In thesis, added E step.

\subsection{Series Evaluation}
Evaluation of the p.d.f.~of the compound Poisson-Gamma distribution is difficult because of the infinite sum
\begin{equation}
p_X(x) = 
\delta(x) \euler^{-\lambda}
+
\euler^{-\beta x-\lambda}\frac{1}{x}\sum_{y=1}^{\infty}W_y
\quad\text{for }x\geqslant 0
\end{equation}
where
\begin{equation}
W_y = \frac{\beta^{y\alpha}\lambda^yx^{y\alpha}}{\Gamma(y\alpha)y!} \ .
\end{equation}
$W_y$ can be expressed in a different way using different parameters
\begin{equation}
W_y(x,p,\phi)=\frac{x^{y\alpha}}{\phi^{y(1+\alpha)}(p-1)^{y\alpha}(2-p)^yy!\Gamma(y\alpha)}
\end{equation}
where
\begin{equation}
p=\frac{2+\alpha}{1+\alpha}
\end{equation}
and
\begin{equation}
\phi = \frac{\alpha+1}{\beta^{2-p}(\lambda\alpha)^{p-1}} \ .
\end{equation}

\cite{dunn2005series} approximated the sum by truncation
\begin{equation}
\sum_{y=1}^\infty W_y \approx \sum_{y=y_\text{l}}^{y_\text{u}}W_y
\end{equation}
where $y_\text{u}<y_{\text{max}}<y_\text{l}$ and $y_{\text{max}}$ is the value of $y$ which maximises $W_y$. \cite{dunn2005series} used Stirling's approximation to find $y_{\text{max}}$. This was done by treating $W_y$ as a continuous function of $y$ and is differentiable with respect to $y$.

@@

The EM algorithm \citep{dempster1977maximum} instead treats $Y\sim\poisson(\lambda)$ as a latent variable. Let $\left\{y_1,y_2,y_3,\dotdotdot, y_n\right\}$ be measurements of $Y$ and $\left\{x_1,x_2,x_3,\dotdotdot, x_n\right\}$ be measurements of $X|Y$. Define the joint log likelihood to be
\begin{multline*}
\ln L(\lambda,\alpha,\beta;X,Y)
=
\sum_{i=1}^n
\left[
	\mathbb{I}(y_i=0)
	\ln
	\prob(Y=0)
	\right.\\\left.
	+
	\mathbb{I}(y_i>0)
	\ln
	\left[
	p_{X|Y}(x_i|y_i)\prob(Y=y_i)
	\right]
\right]
\end{multline*}
so that
\begin{multline}
\ln L(\lambda,\alpha,\beta;X,Y)
=
\sum_{i=1}^n
\left[
	-\mathbb{I}(y_i=0)
	\lambda
	\right.\\\left.+
	\mathbb{I}(y_i>0)
	\left(
	y_i\alpha\ln\beta-\ln\Gamma(y_i\alpha)+(y_i\alpha-1)\ln x_i - \beta x_i
	\right.\right.\\\left.\left.	
	- \lambda + y_i \ln \lambda - \ln(y!)
	\right)
\right] \ .
\end{multline}
The estimators are found by optimising the joint log likelihood. This is done iteratively by estimating the $Y$s given the $X$s and parameters (E step), followed by estimating the parameters given the $Y$s and $X$s (M step) until some convergence conditions are met.

@@

>>>>

\subsection{Series Evaluation}
Evaluation of the p.d.f.~of the compound Poisson-Gamma distribution is difficult because of the infinite sum
\begin{equation}
p_X(x) = 
\delta(x) \euler^{-\lambda}
+
\euler^{-\beta x-\lambda}\frac{1}{x}\sum_{y=1}^{\infty}W_y
\quad\text{for }x\geqslant 0
\end{equation}
where
\begin{equation}
W_y = \frac{\beta^{y\alpha}\lambda^yx^{y\alpha}}{\Gamma(y\alpha)y!} \ .
\label{eq:compoundPoisson_Wy}
\end{equation}
$W_y$ can be expressed in a different way using different parameters
\begin{equation}
W_y(x,p,\phi)=\frac{x^{y\alpha}}{\phi^{y(1+\alpha)}(p-1)^{y\alpha}(2-p)^yy!\Gamma(y\alpha)}
\end{equation}
where
\begin{equation}
p=\frac{2+\alpha}{1+\alpha}
\end{equation}
and
\begin{equation}
\phi = \frac{\alpha+1}{\beta^{2-p}(\lambda\alpha)^{p-1}} \ .
\end{equation}

\cite{dunn2005series} approximated the sum by truncation
\begin{equation}
\sum_{y=1}^\infty W_y \approx \sum_{y=y_\text{l}}^{y_\text{u}}W_y
\end{equation}
where $y_\text{l}<y_{\text{max}}<y_\text{u}$ and $y_{\text{max}}$ is the value of $y$ which maximises $W_y$. \cite{dunn2005series} used Stirling's approximation to find $y_{\text{max}}$. This was done by treating $W_y$ as a continuous function of $y$ and is differentiable with respect to $y$.

@@

The EM algorithm \citep{dempster1977maximum} instead treats $Y\sim\poisson(\lambda)$ as a latent variable. Let $\left\{Y_1,Y_2,Y_3,\dotdotdot, Y_n\right\}$ be realizations of $Y$ and $\left\{x_1,x_2,x_3,\dotdotdot, x_n\right\}$ be measurements of $X|Y$. Define the joint log likelihood to be
\begin{multline*}
\ln L(\lambda,\alpha,\beta;X,Y)
=
\sum_{i=1}^n
\left[
	\mathbb{I}(Y_i=0)
	\ln
	\prob(Y=0)
	\right.\\\left.
	+
	\mathbb{I}(Y_i>0)
	\ln
	\left[
	p_{X|Y}(x_i|Y_i)\prob(Y=Y_i)
	\right]
\right]
\end{multline*}
so that
\begin{multline}
\ln L(\lambda,\alpha,\beta;X,Y)
=
\sum_{i=1}^n
\left[
	-\mathbb{I}(Y_i=0)
	\lambda
	\right.\\\left.+
	\mathbb{I}(Y_i>0)
	\left(
	Y_i\alpha\ln\beta-\ln\Gamma(Y_i\alpha)+(Y_i\alpha-1)\ln x_i - \beta x_i
	\right.\right.\\\left.\left.	
	- \lambda + Y_i \ln \lambda - \ln(Y!)
	\right)
\right] \ .
\end{multline}
The estimators are found by optimising the joint log likelihood. This is done iteratively by estimating the $Y$s given the $X$s and parameters (E step), followed by estimating the parameters given the $Y$s and $X$s (M step) until some convergence conditions are met.

@@

In the E step, the realizations of $Y$ are estimate using
\begin{equation}
y_i = \expectation\left[
Y|X=x_i
\right]
\end{equation}
for given parameters $\lambda$, $\alpha$ and $\beta$. The conditional expectation is calculated using
\begin{equation}
y_i = 
\begin{cases}
0 & \text{ for } x=0 \\ 
\dfrac{\sum_{y=1}^\infty y \prob(Y=y|X=x_i)}{\sum_{y=1}^\infty \prob(Y=y|X=x_i)} & \text{ for } x>0
\end{cases}
\end{equation}
where
\begin{equation*}
\prob(Y=y|X=x) = \frac{p_{X|Y}(x|y)\prob(Y=y)}{p_X(x)} \ .
\end{equation*}
Focussing on the $x>0$ case for now
\begin{equation*}
\prob(Y=y|X=x) = \frac{1}{p_X(x)}\frac{\beta^{y\alpha}}{\Gamma(y\alpha)}x^{y\alpha-1}\euler^{-\beta x}\frac{\euler^{-\lambda}\lambda^y}{y!}
\end{equation*}
and
\begin{equation}
\prob(Y=y|X=x) = W_y \frac{\euler^{-\lambda-\beta x}}{x}
\end{equation}
where $W_y$ is defined in Equation \eqref{eq:compoundPoisson_Wy}.
Then the conditional expectation is
\begin{equation}
y_i = \frac{\sum_{y=1}^\infty y W_y}{\sum_{y=1}^\infty W_y} \ .
\end{equation}
It was already established how the denominator can be evaluated \citep{dunn2005series}. A similar method for evaluating the numerator can be obtained by truncating the sum and summing large terms.

Let
\begin{equation}
W_y^{(r)} = y^r W_y
\end{equation}
so that
\begin{equation}
y_i = \frac{\sum_{y=1}^\infty W^{(1)}_y}{\sum_{y=1}^\infty W_y} \ .
\end{equation}
The exponent $r=0,1,2,\dotdotdot$ will be useful in the M step where higher conditional moments are needed.
The sum can be truncated
\begin{equation}
\sum_{y=1}^\infty W^{(r)}_y \approx \sum_{y=y_\text{l}}^{y_\text{u}} W^{(r)}_y
\end{equation}
where $y_\text{l}<y_\text{max}<y_\text{u}$ and $y_\text{max}$ is the value of $y$ which maximises $W_y^{(r)}$. By expressing $\ln(W^{(r)}_y)$ as
\begin{equation}
\ln\left(W_y^{(r)}\right)=r\ln(y)+\ln(W_y)
\end{equation}
and then taking the derivative with respect to $y$
\begin{equation}
\frac{\partial \ln(W_y^{(r)})}{\partial y} = \frac{r}{y } + \frac{\partial \ln(W_y)}{\partial y} \ .
\end{equation}
Keep in mind that $y=1,2,3,\dotdotdot$ so for large $y$, an approximation can be made $r/y\approx 0$ so that
\begin{equation}
\frac{\partial \ln(W_y^{(r)})}{\partial y} \approx \frac{\partial \ln(W_y)}{\partial y} \ .
\end{equation}
By using such an approximation, the method for evaluating the sum $\sum_{y=1}^\infty W^{(r)}_y$ is almost exactly the same as evaluating the sum $\sum_{y=1}^\infty W_y$, in particular $y_\text{max}$ are the same. For a given $r=0,1,2,\dotdotdot$, the limit of the sum $\sum_{y=y_\text{l}}^{y_\text{u}} W^{(r)}_y$ are chosen such that $W_{y_\text{l}}$ and $W_{y_\text{u}}$ are less than $\epsilon W_{y_\text{max}}^{(r)}$ where $\epsilon$ is some small constant. The limits will be different for different values of $r$.
